---
title: "RNAseq Mouse Kidney"
author: "Uwe Schwartz"
date: "02/12/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_knit$set(root.dir = "~/Analysis/17_20191202_Mm_RNAseq/")
setwd("~/Analysis/17_20191202_Mm_RNAseq/")
out.path<-"~/Analysis/17_20191202_Mm_RNAseq//analysis/"
date.x<-Sys.Date()
```

## Dependencies & Libraries

#### Load packages
```{r libraries, message=FALSE}
library(DESeq2)
library(RColorBrewer)
library(ggplot2)
library(ggrepel)
library(stringr)
library("pheatmap")
palette(unique(c(brewer.pal(9,"Set1"),brewer.pal(8,"Set2"))))
```

##Import data

First load and prepare count table. Subsequently a meta-data table is generated. 

```{r annotation file}
# load count table from featureCounts
count.table<-read.delim("counts/count_table.txt", skip=1)

names1<-sapply(strsplit(colnames(count.table[,7:ncol(count.table)]),
                        split = ".", fixed = T), function(x) 
                            paste(x[(grep("bams", x)+1)]))
colnames(count.table)[7:ncol(count.table)]<-names1
counts<-count.table[,7:ncol(count.table)]
rownames(counts)<-as.character(count.table$Geneid)


## data frame annotation annotation 

replica<-rep(paste0("R",1:6),2)

mouse<-sapply(strsplit(colnames(counts),split = "_"),
                           function(x) x[2])

condition<-sapply(strsplit(colnames(counts),split = "_"),
                           function(x) x[3])

meta.DATA<-data.frame(ID=colnames(counts),replica, 
                      condition,mouse,
                      row.names = colnames(counts))
meta.DATA


## create DESeq obj
dds.obj <- DESeqDataSetFromMatrix(countData = counts,
                                   colData = meta.DATA,
                                   design= ~condition)

```



```{r checkpoint1, eval=FALSE, echo=FALSE }
dir.create(paste0(out.path,"/obj"))
save(dds.obj, file=paste(out.path,"obj/dds.obj.rda",sep="/"))
```


## exploratory analysis

Exploratory analysis was performed to get an overview about the quality of the samples and make replica clustering. First the data is transformed using the rlog function within the DESeq2 package. The function rlog, stands for regularized log, transforming the original count data to the log2 scale by fitting a model with a term for each sample and a prior distribution on the coefficients which is estimated from the data.

```{r PCA, fig.height=5}
# rlog tranformation 
rld<-rlog(dds.obj,blind = T)

## plot PCA

### variables
my.pal<-c(brewer.pal(9,"Set1"))
object<-rld
ntop=500
var.col<-"condition"
var.symb<-"replica"
intgroup=c(var.col,var.symb )
returnData=F
###

rv <- rowVars(assay(object)) 
select <- order(rv, decreasing = TRUE)[seq_len(min(ntop,length(rv)))]

pca <- prcomp(t(assay(object)[select, ]))
percentVar <- pca$sdev^2/sum(pca$sdev^2)


# varaince covered by PC
barplot(percentVar, names.arg = paste0("PCA", 1:length(percentVar)),
          ylab="variance", cex.names = 0.5)


intgroup.df <- as.data.frame(colData(object)[,intgroup,drop = FALSE])
group <- if (length(intgroup) > 1) {
  factor(apply(intgroup.df, 1, paste, collapse = " : "))
} else {
  colData(object)[[intgroup]]
}

# show first two components
i=1
d <- data.frame(PCx = pca$x[, i], PCy = pca$x[, i+1], group = group, 
                  intgroup.df, name=colData(dds.obj)[colnames(object),"ID"])

g<-ggplot(data = d, aes_string(x = "PCx", y = "PCy", color = var.col,
                               shape=var.symb)) +
      geom_point(size = 5) + theme_bw()+xlab(paste0("PC",i," : ",
                                round(percentVar[i] * 100), "% variance"))+
      ylab(paste0("PC",i+1,": ", round(percentVar[i+1] *100),
                  "% variance"))+scale_color_manual(values=my.pal)  
print(g)


```
Samples cluster based on their condition on PC2 which is covering 19% of variance in the data. PC1 might show clustering based on sample collection and preparation with R1 (KO & WT) on the left and R5 (KO & WT) on the right.  

```{r PCA save data, eval=FALSE, echo=FALSE }
dir.create(paste(out.path,"plots_exploratory",sep="/"),
           showWarnings = F)
dir.create(paste(out.path,"plots_exploratory/PCA",sep="/"),
           showWarnings = F)


pdf(paste(out.path,"plots_exploratory/PCA/PCA_overview.pdf",sep="/"), 
    width = 4, height = 4)
  barplot(percentVar, names.arg = paste0("PCA", 1:length(percentVar)),
          ylab="variance", cex.names = 0.5)
dev.off()



for(i in 1:(length(percentVar)-1)){
  d <- data.frame(PCx = pca$x[, i], PCy = pca$x[, i+1], group = group, 
                  intgroup.df, name = colData(dds.obj)[colnames(object),"ID"])
  
pdf(paste0(out.path,"/plots_exploratory/PCA/PC",i,"_and_PC",i+1,".pdf"),
    width=4, height = 4)
    g<-ggplot(data = d, aes_string(x = "PCx", y = "PCy", 
                                   color = var.col, shape=var.symb)) +
      geom_point(size = 3.5) + theme_bw()+
      xlab(paste0("PC",i," : ", round(percentVar[i] * 100), "% variance"))+
      ylab(paste0("PC",i+1,": ", round(percentVar[i+1] *100), "% variance"))+
    scale_color_manual(values=my.pal)  
    print(g)
dev.off()
  
}


```

```{r checkpoint2, eval=FALSE, echo=FALSE }
save(rld, file=paste(out.path,"obj/rld.rda",sep="/"))
```

### heatmaps 

First sample wise similarity clutstering was performed and then the 500 most variable genes were clustered based on their expression levels

```{r heatamaps explorativ, fig.height=5}

## similarity clustering
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)

sampleDists <- dist(t(assay(rld)))

sampleDistMatrix <- as.matrix( sampleDists )
rownames(sampleDistMatrix) <-rld$ID
colnames(sampleDistMatrix) <- NULL

pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors)

## 500 most variable genes

rv <- rowVars(assay(rld))   
select <- order(rv, decreasing = TRUE)[1:500]
mat <-assay(rld)[select,]

anno<-as.data.frame(colData(rld)[,c(var.col,var.symb)])

anno_colors=list()

lv.col<-levels(colData(dds.obj)[,var.col])
anno_colors[[var.col]]<-c(brewer.pal(length(lv.col), "Set1"))[1:length(lv.col)]
names(anno_colors[[var.col]])<-lv.col

lv.symb<-levels(colData(dds.obj)[,var.symb])  
anno_colors[[var.symb]]<-c(brewer.pal(length(lv.symb), "Set2"))
names(anno_colors[[var.symb]])<-lv.symb

pheatmap(mat, annotation_col = anno,scale="row", show_rownames = F,
               annotation_colors =anno_colors)
```

The 500 most variable genes cluster mainly based on a gene set which are specific for R5. A subset of the top variable genes correspond to "WT" and "KO" specific gene signatures (top gene cluster). 

```{r heatmap save plots, eval=FALSE, echo=FALSE }


pdf(paste0(out.path,"/plots_exploratory/similarity_clustering.pdf"), 
           width=5, height=4 ,onefile = F)
  pheatmap(sampleDistMatrix,
         clustering_distance_rows = sampleDists,
         clustering_distance_cols = sampleDists,
         col = colors)
dev.off()



pdf(paste0(out.path,"/plots_exploratory/heatmap_most_var_genes.pdf"),
           height=5, width=5, onefile = F)
    print(pheatmap(mat, annotation_col = anno,scale="row", show_rownames = F,
               annotation_colors =anno_colors))
dev.off()




```




